Effective Inclusion of Electronic Polarization to a Negatively Charged Membrane

### 2019-03-13 Manuscript/manuscript.tex
Update text

\begin{table}[tb!] 
\centering
  \caption{Values of the area per lipid (APL) of POPS bilayers 
	   at 298~K with only \ce{Na^+} counterions and no additional ions. 
           Standard deviations are given. \label{tab:apls} } 
  \begin{tabular}{l|c } 
    \multicolumn{2}{c}{POPS} \\
    model          & APL / Å$^2$    \\ 
    \hline 
    Lipid17/Dang              & 53.5$\pm$ 1.7   \\ 
%    Lipid17/ff99              & 57.9$\pm$ 1.7   \\ 
%    \hline 
    ECC-POPS/ECC-ions         & 59.8$\pm$ 1.8   \\ 
%    \hline 
    experiment (SDP model) \citep{kucerka14} & 62.7$\pm$ 1.3  \\ 
%    \hline 
  \end{tabular} \\
  \todo{I removed the value from Lipid17/ff99 simulation because this was not listed in Table I
    and I think that it is more clear if we refer only to the NMRlipids IV result in the text.
    I also changed the experimental value to correspond the value in the cited paper and added the error bar.}
\end{table}

@@

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[height=\figheight]{../img/ecc_pops/density_profiles_na-k-counterions_wat_phos_compar_purePOPS_ecclipids-lipid17.pdf}
  \includegraphics[height=\figheight]{../img/ecc_pops/density_profiles_na-k-counterions_wat_phos_compar_5PC-1PS_ecclipids-lipid17.pdf}
  \caption{\label{fig:POPS-counterions-dens}
    Number density profiles of \ce{K^{+}}, \ce{Na^{+}} counterions along the membrane normal axis 
    for the negatively charged bilayers composed of only POPS (left) or POPC:POPS (5:1, right) 
    using ECC-lipids/ECC-ions (solid lines) and Lipid17/Dang ions (dashed lines).  
    The density profiles of phosphate groups and water are divided by 4 and 100, respectively.  
}
\todo{JOE: add labels with the relative surface excess $\Gamma$, which are promised to be present in the text.} \\
\todo{JOE: split this figure into two: separate for POPS (to SI) and PC:PS 5:1 (main text)?} \\
\todo{SAMULI: I feel now that we should keep the as it is, but if something is moved to SI, it should be POPS in main text, mixture in SI.}
\end{figure}

@@

First, we monitor the response of POPC headgroup order parameters to the increasing amount of POPS (Fig.~\ref{fig:delta_ordPar_NaCl_PC-PS_mix}).
According to the electrometer concept, the head group order parameters of PC lipids increase
with the increasing amount of negatively charged PS lipids in experiments, because the headgroup P--N dipole
tilts more parallel to the membrane interface \cite{seelig87,scherer87}.
The trend is observed with both force fields in simulations with \ce{K+} counterions and
in ECC-POPS simulation with \ce{Na+} counterions, but not in Lipid17/Dang ion simulations with \ce{Na+} counterions,
where the order parameters are unchanged or decrease and the P--N vector tilts more perpendicular to the membrane interface.
The result can be explained by the higher binding affinity of sodium in the
Lipid17/Dang ion simulations (Fig~\ref{fig:POPS-counterions-dens}), which
compensates the negative charge of PS lipids at the interface.
\begin{figure*}[!tbp] 
  \centering 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPC_nacl.pdf} 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPS_nacl.pdf} 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPC_kcl.pdf} 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPS_kcl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PC-PS_mix} 
    The head group order parameters of a POPC:POPS (5:1) bilayer as a function of PS content
    with \ce{Na^+} (top) and \ce{K^+} (bottom) counterions from simulations with ECC-lipids and Lipid17 \cite{lipid17-future} 
    and experiments (only \ce{Na^+} counterions, but shown also in bottom plots). \cite{roux90}. 
    The underestimated response of ECC-lipids with \ce{Na^+} counterions 
    is likely due to a still slightly overestimated binding affinity of \ce{Na^+} to the phospholipids,
    which is corroborated by the series with \ce{K^+} counterions (lower affinity than \ce{Na^+}),
    where the response is closer to the experiment, which uses only \ce{Na^+} counterions. 
  }
  \todo{JOE+SAMULI: split this figure into separate plots for PC and PS, 
    and change the PS-plot x-axis to reflect the dilution of POPS by PC (now shows the content of PS). 
    The experimental lines in the bottom plot shall be different to make clear that they are not with K counterions (gray dashed?).} \\
  \todo{Error bar of 0.02 should be shown for the experimental POPS results because they come from different experiments, see caption of figure 8 in NMRlipids IV publication.}
\end{figure*}

@@

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[height=\figheight]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPC_nacl.pdf} 
  \includegraphics[height=\figheight]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPS_nacl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PCPS} 
    Changes of the head group order parameters $\alpha$, $\beta$ and the orientations of the carboxylate group and the P-N vector  
    of POPC (left) and POPS (right) phospholipids in a POPC:POPS 5:1 bilayer 
    as a function of \ce{NaCl} concentration in the buffer 
    from simulations with different force fields and experiments at 298 K. \citep{roux90}
    Because data with \ce{NaCl} are not available for POPC, 
    we show experimental data for \ce{LiCl} (dashed line, left) 
    as an upper bound for the magnitude of the response to \ce{NaCl}, 
    which has a lower affinity to phospholipid bilayers compared to \ce{LiCl} \citep{roux90}. 
    The orientation of the \ce{COO^-} group is defined as 
    the connector from the $\beta$ carbon to the carbon in \ce{COO^-} (stars, bottom right). 
  }
  \todo{We should plot the experimental results with KCl also in this figure.} \\
  \todo{Why there are no error bars for ECC-POPS in this and other figures?} \\
  \todo{SAMULI: Use the same y-axis scale in the bottom figures showing angles for POPC and POPS to ease the comparison.}
  \end{figure}

@@

% This§: 
% adsorption and affinities of Na and K to PC:PS memb
% from densities and rel.surf.excess
% 
%in  Fig.~\ref{fig:POPS-counterions-dens} (only counterions)
%and Fig.~\ref{fig:nacl-dens_PCPS}        (additional salt concentrations), 
%
The influence of negatively charged PS on the counterion binding affinity can
be estimated by comparing the relative surface excesses with respect to water, $\Gamma ^{w} _{\rm ion}$,
between POPC:POPS (5:1) mixture and pure POPC \cite{melcr18}.
%Such a quantity compares the adsorption of ions to the adsorption of water molecules at an interface 
%without the necessity of defining a Gibbs dividing surface. \citep{melcr18, chattorajBOOK}
The value of $\Gamma ^{w} _{\rm Na}=-0.11 \pm 0.01$ nm$^{-2}$ for 1 M sodium in pure POPC, calculated from
ECC-POPC simulation, was reported in the previous work \cite{melcr18}. Here, the ECC-lipid simulations
give
%$\Gamma ^{w} _{\rm K}=-0.12 \pm ??$ nm$^{-2}$ and
$\Gamma ^{w} _{\rm Na}=0.09 \pm ??$ nm$^{-2}$ for 0.621 M sodium in pure POPC:POPS (5:1) mixture.
Therefore, the presence of $\sim$17\% POPS in POPC bilayer increases the relative surface
excess of sodium with respect to water from negative to positive. As discussed above,
the sodium binding to POPC:POPS (5:1) mixture may be slightly overestimated also in
ECC-lipid simulations. However, the difference may not be affected because similar small
overestimation seem to occur also in pure POPC simulation where the headgroup order parameter
response to NaCl is also slightly overestimated \cite{meclr18}.
%Interestingly enough, \ce{K^+} maintains negative values of $\Gamma^{w}_{K}$ even for the negatively charged PC:PS~(5:1) bilayer.
%This is also evident from the density plots in  Fig.~\ref{fig:nacl-dens_PCPS}
%showing that the density profile of \ce{K^+} along the membrane normal 
%decays clearly faster towards the membrane centre compared to the density of water. 
%In contrast, the density profile of \ce{Na^+} has a shoulder 
%at the region of the phosphate groups,
%which demonstrates that sodium cations adsorb weakly to the membrane interface compared to the aquaeous solution
%by yielding a small positive value of $\Gamma^{w}_{Na}$.

@@

We have investigated this effect using the mean orientations of P-N and \ce{COO^-} vectors 
(defined and plotted in Fig.~\ref{fig:delta_ordPar_CaCl_PCPS}). 
While the mean angle between the P-N vector of PC head group and the membrane normal
decreased by 11$^\circ$ from 64$^\circ$ to 55$^\circ$
 at 620~mM buffer concentration of \ce{CaCl2},
the mean orientation of the P-N vector in the PS head group has 
decreased only by 3$^\circ$ from 67$^\circ$ to 64$^\circ$ 
in the same concentration range.

Such a largely diminished flexibility 
may arise from the binding of calcium to the \ce{COO^-} group in PS. 
Whe have measured the mean orientation of the \ce{COO^-} group 
as the connector of the carbon atoms 
forming the bond between the group and the $\beta$-carbon of the phospholipid 
(Fig.~\ref{fig:delta_ordPar_CaCl_PCPS}).
The mean angle between the \ce{COO^-} vector and the membrane normal
changes by $11^\circ$ from $62^\circ$ to $73^\circ$ (more towards the membrane interior)
 at 620~mM buffer concentration of \ce{CaCl2}.
The conformational flexibility of the PS head group is hence restrained 
by the binding of calcium cations to the \ce{COO^-} group, 
which reorients towards the membrane interior and to the phosphate region, 
where calcium cations bind predominantly.

Still, the dominant contribution to the binding of calcium comes from the phosphate groups. 
Interestingly enough,
although the \ce{COO^-} groups increase the affinity of calcium towards PS compared to PC,
 there is no extra density at the outmost surface of the membrane. 
This can be understood from the point of the structural rearangements of the \ce{COO^-} group, 
which take place during the interactions with calcium cations
described above and in Fig.~\ref{fig:delta_ordPar_CaCl_PCPS}.

\todo{JOE: suggestion -- what is the charge at the membrane interface of memb+counterions -vs- memb+CaCl2 concnetration?
                         this could relate well to the zeta potential measurements and the Fig 1 in \cite{melcrova16}. }

\begin{table}[tb!] 
\centering
  \caption{%Bulk concentrations, $C _{Ca}$, and buffer concentrations, $C' _{Ca}$, of Ca$^{2+}$;
           %relative surface excess of calcium with respect to water ($\Gamma_{Ca}^{\rm water}$); 
    %and
    Percentages of the population 
           of bound Ca$^{2+}$ to various moieties 
           in a neutral membrane composed of POPC
           and in a negatively charged membrane with the composition 5\,PC:1\,PS.
           \label{tab:binding}} 
  \begin{tabular}{ l | c c } 
        interacting moiety    &  \multicolumn{2}{c}{percentage of bound \ce{Ca^{2+}} } \\
%	\hline
%	$C _{Ca}\,/\,\mathrm{mM}$  &  $240\pm 10 $  &  $280\pm 10 $  \\
%	$C'_{Ca}\,/\,\mathrm{mM}$  &  $400\pm 10 $  &  $350\pm 10 $  \\
%	$\Gamma_{Ca}^{\rm water}\, / \,\mathrm{nm}^{-2}$  &  $0.24 \pm 0.01 $  &  $0.06 \pm 0.01 $  \\
%	\hline
%                             &  \multicolumn{2}{c}{ } \\
                             &  5\,POPC:1\,POPS &  POPC   \\
	\hline
	     PC              &   59   &  100   \\
	     PO$_4$    in PC &   41   &   67   \\
	     carbonyls in PC &   <1   &   ~1   \\
	\hline
	     PS              &    8   &        \\ 
	     PO$_4$  in PS   &    2   &        \\
	     COO$^-$ in PS   &    4   &        \\
	     carbonyls in PS &   <1   &        \\
	\hline
	both PC and PS       &   33   &        \\
  \end{tabular} \\
  \todo{I have commented out the surface excess values (available in text) and concentrations.}
\end{table}

% Description of no. conctacts counting
Such a significantly increased binding also has an impact 
on the structure and populations of the bound calcium cations. 
We have counted contacts between the cations and the oxygen atoms of the lipids
similarly as was done in \cite{melcr18} 
to obtain such a detailed information 
about the interactions of \ce{Ca^{2+}} with various moieties in POPC or POPS. 
The threshold for counting a contact was set to $0.3\,\mathrm{nm}$, 
which encompasses the first peak of the radial distribution function between the cations and the oxygen atoms of the lipids.

% Populations from contacts
The percentages of the populations of membrane-bound calcium cations for various membrane moieties 
are summarized in Table~\ref{tab:binding}.
Even though the negatively charged membrane contains only 18\% of POPS, 
on average 41\% of the total population of bound calcium cations is in contact with PS lipids
with 8\% bound only to them. 
% Where Calcium interacts
While POPC interacts with the calcium cations almost entirely through its phosphate group 
in both neutral \cite{melcr18} and negatively charged membranes, 
the inteaction with POPS lipids happens more through the \ce{COO^-} group than through the \ce{PO4} group. 
Interactions of \ce{Ca^{2+}} with carbonyl groups are also present for both POPC and POPS,
however, they are always accompanied by interactions with phosphate groups. 
This is in contrast to 
simulations with the Berger force field \cite{berger97,mukhopadhyay04}
combined with fluorescent and vibrational sum frequency spectroscopy,
which suggest a significant
calcium binding also to the carbonyls groups \cite{melcrova16}. 
This feature arises likely from the inaccuracy of the Berger model 
and its overestimated binding of cations to that region \cite{nmrlipids_proj4}.

Measurements using infrared spectroscopy 
combined with simulations with the CHARMM36 model 
with a correction for overestimated binding of calcium \cite{valentine18}
reveal similar trends for the populations at various PS moieties
suggesting that POPS interacts mostly through its \ce{COO^-} group. 
While this feature agrees nicely with our findings, 
the balance of the interactions of PS moieties with calcium cations may be suboptimal
and the interaction strength underestimated in total in those simulations \cite{nmrlipids_proj4}.

In contrast, the combination of NMR experiments 
and MD simulations with the same model, CHARMM36, but without the correction for calcium binding
also reveal long-lived interactions of \ce{Ca^{2+}} with \ce{PO4} groups. \cite{hallock18}
Although the ratios of the life times of coordination follow the same trends as in our work,
the lengths of simulation trajectories are insufficient 
to yield converged results in this aspect (see SI and Refs.~\cite{catte16, melcr18, nmrlipids_proj4}). 
Moreover, the content of POPS in the bilayers in that work is
much higher above the threshold after which 
\ce{Ca^{2+}} cations induce lipid clustering. \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}
At such a high content of negatively charged lipids,
PS head groups are found to form even four-membered clusters \cite{hallock18}. 
Our simulations are, however, at different conditions,
in which the clustering of POPS does not happen, 
and PS lipids prefer 1:1 ratio with \ce{Ca^{2+}}
as plotted in Fig.~\ref{fig:cacl_complexes}
showing relative probabilities of \ce{Ca^{2+}} complexes with a certain number of PC or PS lipids. 
Compared to a neutral POPC bilayer
the negatively charged bilayer has its stoichiometry 
shifted towards complexes with three phospholipids and one calcium. 
This is obviously due to the presence of POPS, 
which also contributes to the pre-formed one- and two-membered clusters of POPC. 
Clusters of four or more lipids were not observed in either membrane.

\todo{ THE REMAINING PIECES TO BE DISCUSSED  --  consider wheter the dynamics slowdown or the dihedral distributions could make good contributions to this paper. }

% Ca2+ induces much slower dynamics of lipids when PS is present,
% this does not happen with PC
% uses NBFIX
From Ref.~\cite{valentine18}://
To summarize, calcium binding to PS lipids slows the dynamics at the ester position and introduces more heterogeneous molecular environments. 
The slowdown cannot be attributed to complete ester dehydration or decreased hydrogen bonding to water, so we interpret these results as evidence that the lipid headgroups adopt more rigid but disordered conformations on the molecular level. 
The decrease in dynamics observed experimentally indicates that the network of hydrogen bonds between lipids and water at the polar-nonpolar interface becomes slower, as it was shown by Karathanou and Bondar (34). 
This is bolstered by recent NMR work showing that binding to Ca2+ induces two different rigid conformations in the glycerol backbone of lipid headgroups (18).
We have investigated the molecular origin of the rigidification and decreased lateral diffusion, finding that calcium binding does not dehydrate the ester region, but even with only 10\% PS lipids, the local dynamics slow around the esters. The slower local dynamics provide a molecular explanation for the Ca2+-dependent decreases in fluidity observed on longer length and timescales.
There is considerable room for further improvement in the force fields for ion-membrane interactions. Particularly, small differences in short-range interactions may determine the local geometries (34
). Most of our conclusions are, however, based on long-range effects (i.e., electric field generated by the ion at significant distances from the carbonyl groups), which are less sensitive to the details of the force field.

>>>>

\begin{table}[tb!] 
\centering
  \caption{Area per lipid (APL) of POPS bilayers from simulations with different force fields at 298~K with \ce{Na^+} counterions. \label{tab:apls} } 
  \begin{tabular}{l|c } 
    \multicolumn{2}{c}{POPS} \\
    model          & APL / Å$^2$    \\ 
    \hline 
    Lipid17/Dang              & 53.5$\pm$ 1.7   \\ 
%    Lipid17/ff99              & 57.9$\pm$ 1.7   \\ 
%    \hline 
    ECC-POPS/ECC-ions         & 59.8$\pm$ 1.8   \\ 
%    \hline 
    experiment (SDP model) \citep{kucerka14} & 62.7$\pm$ 1.3  \\ 
%    \hline 
  \end{tabular} \\
  \todo{I removed the value from Lipid17/ff99 simulation because this was not listed in Table I
    and I think that it is more clear if we refer only to the NMRlipids IV result in the text.
    I also changed the experimental value to correspond the value in the cited paper and added the error bar.}
\end{table}

@@

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[height=\figheight]{../img/ecc_pops/density_profiles_na-k-counterions_wat_phos_compar_purePOPS_ecclipids-lipid17.pdf}
  \includegraphics[height=\figheight]{../img/ecc_pops/density_profiles_na-k-counterions_wat_phos_compar_5PC-1PS_ecclipids-lipid17.pdf}
  \caption{\label{fig:POPS-counterions-dens}
    Number density profiles of \ce{K^{+}}, \ce{Na^{+}} counterions along the membrane normal axis 
    for the negatively charged bilayers composed of only POPS (left) or POPC:POPS (5:1, right) 
    using ECC-lipids/ECC-ions (solid lines) and Lipid17/Dang ions (dashed lines).  
    The density profiles of phosphate groups and water are divided by 4 and 100, respectively.  
}
\todo{JOE: add labels with the relative surface excess $\Gamma$, which are promised to be present in the text.
  SAMULI: I think that these are not promised in the current text and addition is not necessary.} \\
\todo{JOE: split this figure into two: separate for POPS (to SI) and PC:PS 5:1 (main text)?
  SAMULI: I feel now that we should keep the figure as it is. Also the POPC:POPS (5:1) results are now referred in the text.
  However, if something is moved to SI, it should be POPS in main text, mixture in SI.}
\end{figure}

@@

First, we monitor the response of POPC headgroup order parameters to the increasing amount of POPS (Fig.~\ref{fig:delta_ordPar_NaCl_PC-PS_mix}).
According to the electrometer concept, the head group order parameters of PC lipids increase
with the increasing amount of negatively charged PS lipids in experiments, because the headgroup P--N dipole
tilts more parallel to the membrane interface \cite{seelig87,scherer87}.
The trend is observed with both force fields in simulations with \ce{K+} counterions and
in ECC-POPS simulation with \ce{Na+} counterions, but not in Lipid17/Dang ion simulations with \ce{Na+} counterions,
where the order parameters are unchanged or decrease and the P--N vector tilts more perpendicular to the membrane interface.
The result can be explained by the higher binding affinity of sodium in the
Lipid17/Dang ion simulations (Fig~\ref{fig:POPS-counterions-dens}), which
compensates the negative charge of PS lipids at the interface.
\begin{figure*}[!tbp] 
  \centering 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPC_nacl.pdf} 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPS_nacl.pdf} 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPC_kcl.pdf} 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPS_kcl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PC-PS_mix} 
    CAPTION FOR THE FIGURE WITH POPC: The POPC head group order parameters and the P--N vector angle
    with respect to the membrane normal as a function of POPS content in a bilayer
    from ECC-lipid and Lipid17/Dang simulations with \ce{Na^+} (top) and \ce{K^+} (bottom) counterions.
    Experimental order parameter values are from Ref. \citenum{scherer87}
    and the signs from Ref. \citenum{ferreira16} (only \ce{Na^+} counterions, but shown also in bottom plots).
    CAPTION FOR THE FIGURE WITH POPS: The POPS head group order parameters, the P--N and C$_\beta$--C$_\gamma$ vector angles (stars)
    with respect to the membrane normal as a function of POPC content in a bilayer
    from ECC-lipid and Lipid17/Dang simulations with \ce{Na^+} (top) and \ce{K^+} (bottom) counterions.
    Experimental order parameter values and signs for pure POPS are from Ref. \citenum{nmrlipids_proj4} and values
    for POPC:POPS (5:1) mixture from Ref. \citenum{roux90} (only \ce{Na^+} counterions, but shown also in bottom plots).
    %The underestimated response of ECC-lipids with \ce{Na^+} counterions 
    %is likely due to a still slightly overestimated binding affinity of \ce{Na^+} to the phospholipids,
    %which is corroborated by the series with \ce{K^+} counterions (lower affinity than \ce{Na^+}),
    %where the response is closer to the experiment, which uses only \ce{Na^+} counterions. 
  }
  \todo{JOE+SAMULI: split this figure into separate plots for PC and PS, 
    and change the PS-plot x-axis to reflect the dilution of POPS by PC (now shows the content of PS). 
    The experimental lines in the bottom plot shall be different to make clear that they are not with K counterions (gray dashed?).} \\
  \todo{Error bar of 0.02 should be shown for the experimental POPS results because they come from different experiments, see caption of figure 8 in NMRlipids IV publication.}
\end{figure*}

@@

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[height=\figheight]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPC_nacl.pdf} 
  \includegraphics[height=\figheight]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPS_nacl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PCPS} 
    Changes of the head group order parameters, the P--N and C$_\beta$--C$_\gamma$ vector angles (stars)
    with respect to the membrane normal of POPC (left) and POPS (right) in a POPC:POPS (5:1) bilayer 
    as a function of \ce{NaCl} concentration from ECC-lipid and Lipid17/Dang simulations at 298 K \cite{roux90}.
    Because data with \ce{NaCl} are not available for POPC, 
    we show experimental data for \ce{LiCl} (dashed line, left) 
    as an upper bound for the magnitude of the response to \ce{NaCl}, 
    which has a lower affinity to phospholipid bilayers compared to \ce{LiCl} \citep{roux90}. 
    %The orientation of the \ce{COO^-} group is defined as 
    %the connector from the $\beta$ carbon to the carbon in \ce{COO^-} (stars, bottom right). 
  }
  \todo{We should plot the experimental results with KCl also in this figure.} \\
  \todo{Why there are no error bars for ECC-POPS in this and other figures?} \\
  \todo{SAMULI: Use the same y-axis scale in the bottom figures showing angles for POPC and POPS to ease the comparison.}
  \end{figure}

@@

% This§: 
% adsorption and affinities of Na and K to PC:PS memb
% from densities and rel.surf.excess
% 
%in  Fig.~\ref{fig:POPS-counterions-dens} (only counterions)
%and Fig.~\ref{fig:nacl-dens_PCPS}        (additional salt concentrations), 
%
The influence of negatively charged PS on the counterion binding affinity can
be estimated by comparing the relative surface excesses with respect to water, $\Gamma ^{w} _{\rm ion}$,
between POPC:POPS (5:1) mixture and pure POPC \cite{melcr18}.
%Such a quantity compares the adsorption of ions to the adsorption of water molecules at an interface 
%without the necessity of defining a Gibbs dividing surface. \citep{melcr18, chattorajBOOK}
The value of $\Gamma ^{w} _{\rm Na}=-0.11 \pm 0.01$ nm$^{-2}$ for 1 M sodium in pure POPC, calculated from
ECC-POPC simulation, was reported in the previous work \cite{melcr18}. Here, the ECC-lipid simulations
give
%$\Gamma ^{w} _{\rm K}=-0.12 \pm ??$ nm$^{-2}$ and
$\Gamma ^{w} _{\rm Na}=0.09 \pm ??$ nm$^{-2}$ \todo{Error bar should be added.} for 0.621 M sodium in pure POPC:POPS (5:1) mixture.
Therefore, the presence of $\sim$17\% POPS in POPC bilayer increases the relative surface
excess of sodium with respect to water from negative to positive. As discussed above,
the sodium binding to POPC:POPS (5:1) mixture may be slightly overestimated also in
ECC-lipid simulations. However, the difference may not be affected because similar small
overestimation seem to occur also in pure POPC simulation where the headgroup order parameter
response to NaCl is also slightly overestimated \cite{meclr18}.
%Interestingly enough, \ce{K^+} maintains negative values of $\Gamma^{w}_{K}$ even for the negatively charged PC:PS~(5:1) bilayer.
%This is also evident from the density plots in  Fig.~\ref{fig:nacl-dens_PCPS}
%showing that the density profile of \ce{K^+} along the membrane normal 
%decays clearly faster towards the membrane centre compared to the density of water. 
%In contrast, the density profile of \ce{Na^+} has a shoulder 
%at the region of the phosphate groups,
%which demonstrates that sodium cations adsorb weakly to the membrane interface compared to the aquaeous solution
%by yielding a small positive value of $\Gamma^{w}_{Na}$.

@@

% Description of no. conctacts counting
%Such a significantly increased binding also has an impact 
%on the structure and populations of the bound calcium cations. 
%We have counted contacts between the cations and the oxygen atoms of the lipids
%similarly as was done in \cite{melcr18} 
%to obtain such a detailed information 
%about the interactions of \ce{Ca^{2+}} with various moieties in POPC or POPS.

% Populations from contacts
Interaction sites between calcium and PS lipids \cite{??} have been recently analyzed using
different force fields and experimental techniques, but the results have been controversial.
Simulations using the NBfix parameters for calcium \cite{kim16} in CHARMM36 force field \cite{klauda10,venable13},
combined with 2D infrared spectroscopy, suggest that calcium ions interact only with the carboxylate group of PS lipids \cite{valentine18}.
However, the same force field without the NBfix parameters, combined with NMR chemical shifts and
rotational-echo double-resonance (REDOR) experiments, suggests a significant binding affinity also to the phosphate region \cite{hallock18}.
On the other hand, simulations with the Berger force field \cite{berger97,mukhopadhyay04},
combined with fluorescent and vibrational sum frequency spectroscopy, suggest a significant
calcium binding also to the carbonyls in the acyl chains \cite{melcrova16}.
The advantage of the NMR order parameters used in this work is the direct connection between
measured and simulation data, which minimizes the human interpretation when evaluating the
force field quality \cite{ollila16}. Therefore, we believe that the ECC-lipid model, which is in the best
agreement with experimental headgroup order parameters among the available models, gives the currently best available
interpretation of the experimental data.

In ECC-lipid simulation of POPC:POPS (5:1) mixture with 409 mM CaCl$_2$,
the calcium ions bind twice more likely to the carboxylate than to phosphate moiety of POPS headgroup
\todo{I am not sure if this is correct, because I do not understand the numbers in the table.} (Table \ref{tab:binding}),
which is in line with the results with NBfix correction in CHARMM36 \cite{hallock18}.
However, calcium ions coordinate with even four distinct PS lipids
in CHARMM36 simulations without NBfix \cite{hallock18}, while coordination with more than two PS lipids
was not observed in our ECC-lipid simulations (Fig.~\ref{fig:cacl_complexes}).
This is not surprising because calcium significantly overbinds to PC lipids in
CHARMM36 simulations without the NBfix correction for calcium \cite{catte16}.
Furthermore, the POPS:POPC:Ca$^{2+}$ (7:3:7) mixture with relatively high
PS lipid fraction and calcium concentration was used in previous work \cite{hallock18},
which may lead to the complexation of \ce{Ca^{2+}} cations and PS lipids that is
not observed with lower PS content used in ECC-lipid simulations in this work \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
%While POPC interacts with the calcium cations almost entirely through its phosphate group 
%in both neutral \cite{melcr18} and negatively charged membranes, 
%the inteaction with POPS lipids happens more through the \ce{COO^-} group than through the \ce{PO4} group. 
%Interactions of \ce{Ca^{2+}} with carbonyl groups are also present for both POPC and POPS,
%however, they are always accompanied by interactions with phosphate groups. 
Significant coordination of calcium with carbonyls groups of PS and PC lipids observed in previous
simulations with the Berger force field \cite{melcrova16} %\cite{berger97,mukhopadhyay04}
was not observed in ECC-lipid simulations in here nor in the previous work with pure POPC \cite{melcr18}.
%This is in contrast to 
%combined with fluorescent and vibrational sum frequency spectroscopy,
%which suggest a 
Overestimated coordination of cations with carbonyl groups is a potential source for
overestimated cation binding in Berger force fields \cite{catte16,nmrlipids_proj4}. 
In CHARMM36 simulations with the NBfix, calcium mainly interacted with only
carboxylate of PS lipids  \cite{valentine18}, which is a potential source for the
underestimated calcium binding affinity to lipid bilayer with this force field \cite{nmrlipids_proj4}.

%In contrast, the combination of NMR experiments 
%and MD simulations with the same model, CHARMM36, but without the correction for calcium binding
%also reveal long-lived interactions of \ce{Ca^{2+}} with \ce{PO4} groups. \cite{hallock18}
%Although the ratios of the life times of coordination follow the same trends as in our work,
%the lengths of simulation trajectories are insufficient 
%to yield converged results in this aspect (see SI and Refs.~\cite{catte16, melcr18, nmrlipids_proj4}). 
\begin{table}[tb!] 
\centering
\caption{%Bulk concentrations, $C _{Ca}$, and buffer concentrations, $C' _{Ca}$, of Ca$^{2+}$;
  % relative surface excess of calcium with respect to water ($\Gamma_{Ca}^{\rm water}$); 
  % and
  Percentages of the population of bound Ca$^{2+}$ to various lipid moieties in a pure POPC bilayer with 350 mM CaCl$_2$
  and in a POPC:POPS (5:1) mixture with 409 mM CaCl$_2$. The threshold for counting a contact was set to $0.3\,\mathrm{nm}$, which encompasses the
  first peak of the radial distribution function between the cations and the oxygen atoms of the lipids. 
  \label{tab:binding}} 
\begin{tabular}{ l | c c } 
        interacting moiety    &  \multicolumn{2}{c}{percentage of bound \ce{Ca^{2+}} } \\
%	\hline
%	$C _{Ca}\,/\,\mathrm{mM}$  &  $240\pm 10 $  &  $280\pm 10 $  \\
%	$C'_{Ca}\,/\,\mathrm{mM}$  &  $400\pm 10 $  &  $350\pm 10 $  \\
%	$\Gamma_{Ca}^{\rm water}\, / \,\mathrm{nm}^{-2}$  &  $0.24 \pm 0.01 $  &  $0.06 \pm 0.01 $  \\
%	\hline
%                             &  \multicolumn{2}{c}{ } \\
                             &  5\,POPC:1\,POPS &  POPC   \\
	\hline
	     PC              &   59   &  100   \\
	     PO$_4$    in PC &   41   &   67   \\
	     carbonyls in PC &   <1   &   ~1   \\
	\hline
	     PS              &    8   &        \\ 
	     PO$_4$  in PS   &    2   &        \\
	     COO$^-$ in PS   &    4   &        \\
	     carbonyls in PS &   <1   &        \\
	\hline
	both PC and PS       &   33   &        \\
  \end{tabular} \\
  \todo{I have commented out the surface excess values (available in text) and concentrations (now in caption) from the table.} \\
  \todo{I do not understand the numbers for different moieties. Are these probabilities that the calcium binds only to these moieties and
    nothing else? In total, 41 \% binds to PS lipids, but only 2 \% and 4 \% to phosphates and COO$^-$, respectively? Where do the rest bind?
    I think that we should show the total probabilities to bind to different moieties.}
\end{table}

\begin{figure}[tb!] 
  \centering 
  \includegraphics[width=\figwidth]{../img/stoichiometry_CaCl2_comparison_Ecc-lipids_PC-vs-PCPS.pdf} \\ 
  \caption{\label{fig:cacl_complexes} 
    Relative probabilities of \ce{Ca^{2+}} ions to coordinate with a certain number of lipids
    in pure POPC bilayer with 350 mM CaCl$_2$ and POPC:POPS (5:1) mixture with 400 mM CaCl$_2$.  
    Analysis was done for both systems by considering all lipids (blue and violet) and
    for POPC:POPS (5:1) mixture also by considering POPC and POPS lipids separately (green and orange).
    %All lipids were taken into account with the exception of the complexes in light green and orange, 
    %for which we counted only contacts with POPC resp POPS from the mixed 5\,PC:1\,PS negatively charged bilayer. 
    %The calculated probabilities of the calcium-lipid complexes also reflect only POPC (light green) resp POPS (orange). 
    % Probabilities were taken from simulations with comparable bulk concentrations of calcium around 250~mM. 
    Clusters of four or more lipids were not observed in either membrane.
    Previously published simulation data \cite{melcr18} for pure POPC bilayers were taken directly from ?? Add Zenodo reference, if available. 
  }
  \todo{Add definition used for coordination and link to the Zenodo, if available.}
\end{figure}
